Four decades of water quality change in the upper San Francisco Estuary\footnote{Version: Fri Jan 13 16:19:14 2017 -0600, \href{https://github.com/fawda123/sftrends_manu/commit/66d301f76d51fddfc7d4eeea2ed7b92995d7385f}{66d301f76d51fddfc7d4eeea2ed7b92995d7385f}}

### 2017-01-15 sftrends_manu.Rnw
trndmap gone, almost done with trends text

Daily flow estimates for the Delta region were obtained from the Dayflow software program that provides estimates of average Delta outflow \cite{IEP16}.  Because of the complexity of water inflow, exports, and outflows from the Delta, the Dayflow program combines observations with estimates based on mass balance to reconstruct historical and daily flow estimates.  The \ac{wrtds} models described below require a matched flow record with the appropriate station to evaluate nutrient trends. Given the complexity of inflows and connectivity of the system, only the inflow estimates from the Sacramento and San Joaquin rivers were used as measures of freshwater influence at each station.  Initial analyses indicated that model fit was not significantly improved with flow estimates from locations closer to each station, nor was model fit improved using lagged times series.  As such, the Sacramento daily flow time series was used to acount for flow effects at C3, D19, D26, and D28, and the San Joaquin time series was used for C10 and P8.  The salinity observations at D4, D6, and D7 in Suisun Bay were used as a more appropriate measure of variation in freshwater balance given the stronger tidal influence at these stations.  Salinity has been used as a tracer of freshwater influence for the application of WRTDS models in tidal waters \cite{Beck15}.

@@

Relationships between flow and observed water quality are complex and can change significantly through space and time\cite{Hirsch10,Zhang16}. These principles have been demonstrated for monitoring data in the Delta region\cite{Jassby08,Novick15,Jabusch16}, suggesting that trend analyses using the observed time series are confounded by flow effects.  Change over for time in the observed reflect could reflect mobilization or dilution effects of flow on concentration rather than an empirical system response to changes in nutrient sources.  As a proof of concept, \cref{fig:dinc10} demonstrates use of \ac{wrtds} to isolate a flow-normalized time series from the observed data. Raw data are presented in \cref{fig:dinc10}a and the annual results by water year (October through September) from \ac{wrtds} are shown in \cref{fig:dinc10}b. In addition to removing the seasonal component, \cref{fig:dinc10}b shows the flow-normalized component (solid line) independent of the model predictions.  The difference between the two is shown in \cref{fig:dinc10}c such that years with predictions greater or less than the flow-normalized values correspond with long-term trends in flow shown in \cref{fig:dinc10}d.  For example, 1984 is a period of high flow and a large, negative difference between prediction and flow-normalized concentration, suggesting a dilution effect of increased flow on nutrient concentration.  Further, \cref{fig:dinc10}e shows seasonal variation in the relationships of \ac{din} with flow throughout the period of record. Increases in flow (y-axis) were associated with an increase in \ac{din} (colors) for flow values within the observed range.  Seasonal patterns also differed througout the time period with a wider range of \ac{din} within a growing season in the early 2000s relative to the 1980s. % The seasonal shifts are potentially linked to seasonal upwelling linked to ENSO \cite{Cloern07}

@@

zcmp <- select(cmp, -perchg) %>% 
  mutate(ztest = ifelse(ztest < .05, 'sig', 'ns')) %>% 
  spread(dat, ztest) %>% 
  mutate(
    sigtons = ifelse(obs == 'sig' & nrm == 'ns', 1, 0), 
    nstosig = ifelse(obs == 'ns' & nrm == 'sig', 1, 0)
  )

chgcmp <- select(cmp, -ztest)# %>% 
  # mutate(ztest = ifelse(ztest < .05, 'sig', 'ns')) %>% 
  # spread(dat, ztest) %>% 
  # mutate(
  #   sigtons = ifelse(obs == 'sig' & nrm == 'ns', 1, 0), 
  #   nstosig = ifelse(obs == 'ns' & nrm == 'sig', 1, 0)
  # )

@@

A comparison of trends with flow-normalized results from \ac{wrtds} relative to observed data is justified because flow and nutrient concentrations were linked at many of the stations in the study area, similar to \cref{fig:dinc10}.  These comparisons are made relative to changes in the magnitude, significance, and direction of trends, all of which have important implications for decision-making. For all \Sexpr{eng(nrow(zcmp))} trend comparisons in \cref{fig:trndcomp1} regardless of site, nitrogen analyte, and time period, \Sexpr{eng(sum(zcmp$nstosig))} comparisons had trends that were insignificant with the observed data but significant with flow-normalized results. The magnitude of all comparisons changed

Overall, the differences in apparent trends underscore the critical importance of considering flow effects in the interpretation of environmental changes, where flow in the Delta is affected by complex relationships between river inputs, water withdrawals, and tidal exchange with the larger Bay.

@@

Implications of different flow varialbes - refer to \cite{Novick15}, a hybrid approach would be best for complex systems like the delta.

@@

<<trndmap, fig.cap = 'Percent change per year in nitrogen analytes for aggregations by seasons from (a) 1976-1995 and (b) 1996-2013.  Changes are based on seasonal Kendall tests of flow-normalized results within each time period. Station names are shown in the top left panel. Station locations have been jittered to reduce overlap. Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.', fig.height = 8, fig.width = 5.3, cache = T, out.width = '0.7\\textwidth'>>=

# load required data
data(delt_dat)
data(delt_map)
data(trnds_seasyr)

# combine all
trnds_seasyr <- rbind(trnds_nrm, trnds_fit, trnds_obs) %>% 
  select(Site_Code, resvar, cat, perchg, ann, trndvar) %>% 
  spread(trndvar, perchg) %>% 
  mutate(Site_Code = factor(Site_Code, levels = c('D7', 'D6', 'D4', 'D28', 'D26', 'D19', 'MD10', 'P8', 'C10', 'C3')))

##
# get a bounding box for the stations
statmeta <- select(delt_dat, Site_Code, Latitude, Longitude) %>% 
  unique

# bounding box
buffx <- 0.00035
buffy <- 0.0022
lims<- select(statmeta, -Site_Code) %>% 
  apply(., 2, function(x) range(x, na.rm = TRUE))
lims <- c(
  (1 + buffx) * lims[1, 2], 
  (1 - buffy) * lims[1, 1], 
  (1 - buffx) * lims[2, 2], 
  (1 + buffy) * lims[2, 1]
  )    
names(lims) <- c('left', 'bottom', 'right', 'top')

# need to shift locations of some stations
statmeta[statmeta$Site_Code %in% 'D26', 'Latitude'] <- 38.14
statmeta[statmeta$Site_Code %in% 'D26', 'Longitude'] <--121.55
statmeta[statmeta$Site_Code %in% 'D7', 'Latitude'] <- 38.13
statmeta[statmeta$Site_Code %in% 'D7', 'Longitude'] <--122.015
statmeta[statmeta$Site_Code %in% 'C10', 'Latitude'] <-37.68
statmeta[statmeta$Site_Code %in% 'C10', 'Longitude'] <--121.28
statmeta[statmeta$Site_Code %in% 'D28', 'Latitude'] <-37.94
statmeta[statmeta$Site_Code %in% 'MD10', 'Latitude'] <-38.1
statmeta[statmeta$Site_Code %in% 'P8', 'Latitude'] <-37.93
statmeta[statmeta$Site_Code %in% 'D19', 'Latitude'] <-38.06

# response labels 
reslabs <- c('din an', 'nh an', 'no23 an', 'din mobef', 'nh mobef', 'no23 mobef', 'din moaft', 'nh moaft', 'no23 moaft')
reslevs <- rep(c(
  'DIN', 
  expression(paste(NH [4] ^ '+')),
  expression(paste(NO [2] ^ '-', ' / ', NO [3] ^ '2-'))
  ), 3)

# convert delta map to ggplot compatible object
delt_map <- fortify(delt_map)

# base delta map
alph <- 1
pbase <- ggplot(delt_map, aes(x = long, y = lat)) + 
  geom_polygon(aes(group = group, fill = hole), colour = alpha("lightblue", alph)) +
  theme_bw() + 
  theme(
    plot.background = element_rect(fill='transparent', 
      colour = NA),
    panel.background = element_rect(fill='transparent', 
      colour = NA),
    legend.background = element_rect(fill='transparent', 
      colour = NA),
    legend.key = element_rect(fill = 'transparent', 
      colour = NA),
    axis.ticks.x = element_line(),
    axis.ticks.y = element_line(),
    axis.ticks.length = unit(.1, "cm"), 
    panel.grid.minor = element_blank(), 
    axis.title.x = element_blank(), 
    axis.title.y = element_blank(), 
    legend.position = 'none', 
    plot.margin = grid::unit(c(3, 3, 3, 3), 'pt'), 
    axis.text.x = element_text(size = 8), 
    axis.text.y = element_text(size = 8), 
    strip.background=element_rect(fill = 'white', colour = 'white'),
    panel.border = element_rect(colour = "grey")
  ) +
  coord_fixed(ratio = 1, xlim = lims[c(1, 3)], ylim = lims[c(2, 4)]) +
  scale_fill_manual(values=c(alpha("lightblue", alph), "white"), guide="none")

# bounds on limits
yrng <- max(abs(range(trnds_seasyr$norm)))
yrng <- c(-1 * yrng, yrng)

# color ramp, red for pos, green for neg
colpos <- RColorBrewer::brewer.pal(9, 'Reds')[3:9]
colneg <- RColorBrewer::brewer.pal(9, 'Greens')[3:9]
palneg <-  colorRampPalette(colneg)(1000)
palpos <-   colorRampPalette(colpos)(1000)

# plot(1:2000, col = c(palneg, palpos), pch = 16, cex = 4)
# points(1000, 1000)

# unnest data frame, assign colors by rank and sign
toplo <- select(trnds_seasyr, Site_Code, resvar, ann, cat, norm) %>% 
  rename(perchg = norm) %>% 
  left_join(., statmeta, by = 'Site_Code') %>% 
  data.frame %>% 
  mutate(
    perchg = pmax(yrng[1], perchg),
    perchg = pmin(yrng[2], perchg),
    negs = sign(perchg),  
    rnks = scales::rescale(abs(perchg), c(1, 1000)), 
    rnks = round(rnks), 
    cols = ifelse(negs == 1, palpos[rnks], palneg[rnks]),
    ind = 1:nrow(.)
    ) %>%
  filter(!cat %in% c('1976-1995', '1996-2013')) %>% # remove year categories
  unite('cattype', cat, ann, sep = '', remove = F) %>%
  mutate(
    cattype = factor(cattype),
    cattype = fct_collapse(cattype, 
      mobef = paste0(c('Spring', 'Summer', 'Fall', 'Winter'), 'bef'), 
      moaft = paste0(c('Spring', 'Summer', 'Fall', 'Winter'), 'aft') 
    )
  )

# unite response varibles and summary period combos
unis <- with(toplo, paste(resvar, cattype)) %>% 
  unique %>% 
  sort

sites <- unique(toplo$Site_Code)

plo_ls <- vector('list', length = length(unis))
names(plo_ls) <- unis

for(uni in unis){

  # get variable, agg period
  splits <- strsplit(uni, ' ')[[1]]
  resvart <- splits[1]
  cattypet <- splits[2]
  toplo_tmp <- filter(toplo, resvar %in% resvart & cattype %in% cattypet)

  # create individual bar plots as ggplot list
  bars <- setNames(lapply(sites, function(i){

    subplo <- filter(toplo_tmp, Site_Code %in% i)
    ggplot(subplo, aes(x=cat, y = perchg, fill = factor(cat))) +
      geom_bar(stat = 'identity', colour = 'black', size = 0.2) + 
      scale_fill_manual(values = as.character(subplo$cols)) + 
      theme_tree() + 
      xlab(NULL) + 
      ylab(NULL) + 
      theme_transparent() +
      theme(plot.margin=unit(c(0,0,0,0),"mm")) + 
      scale_y_continuous(limits = yrng)

  }), sites)

  ##
  # add the barplots to the base map

  # add site labels if plot is first
  pall <- pbase
  set.seed(5)
  if(which(uni == unis) == 2)
    pall <- pbase + 
      geom_label_repel(
        data = statmeta, 
        aes(x = Longitude, y = Latitude, label = Site_Code), 
        label.r = unit(0, "lines"),
        box.padding = unit(2, "lines"),
        label.padding = unit(0.15, "lines"), 
        point.padding = unit(0.1, "lines"),
        force = 2, size = 2,
        fill = 'grey95'
      )

  # keep bar widths the same for cattypes
  subwid <- 0.1
  if(grepl('an$', uni))
    subwid <- 0.06

  for(i in 1:length(bars)){

      nms <- names(bars)[i]
      dat <- statmeta[statmeta$Site_Code %in% nms,]
      pall <- subview(pall, bars[[i]], x = dat$Longitude, y = dat$Latitude, subwid, 0.21) # change bar heights here

  }

  tval <- reslevs[which(uni == reslabs)]
  pall <- pall + 
    geom_text(data = data.frame(x = lims[1], y = lims[4], tval = as.character(tval)), 
      aes(x = x, y = y, label = tval), parse = TRUE, hjust = 0, vjust = 1)
  plo_ls[[uni]] <- pall

}

# setup legend
rngs <- seq(yrng[1], yrng[2], length = length(c(colneg, colpos)))
leg <- ggplot(data.frame(x = seq_along(rngs), y = seq_along(rngs), z = rngs), aes(x, y, colour = z)) + 
  geom_point() + 
  scale_colour_gradientn(colours = c(rev(colneg), colpos)) + 
  guides(colour = guide_colourbar(title = expression('% chg yr' ^-1), barheight = 0.5, barwidth = 8)) + 
  theme(legend.position = 'top')
leg <- g_legend(leg)

grid.arrange(
  leg,
  arrangeGrob(
    arrangeGrob(plo_ls[[2]], plo_ls[[4]], plo_ls[[6]], ncol = 1, top = c('(a) Sp, Su, Fa, Wi (1976-1995)')),
    arrangeGrob(plo_ls[[1]], plo_ls[[3]], plo_ls[[5]], ncol = 1, top = c('(b) Sp, Su, Fa, Wi (1996-2013)')),
    ncol = 2
    ),
  ncol = 1, 
  heights = c(0.1, 1)
  )

>>>>

Daily flow estimates for the Delta region were obtained from the Dayflow software program that provides estimates of average Delta outflow \cite{IEP16}.  Because of the complexity of water inflow, exports, and outflows from the Delta, the Dayflow program combines observations with estimates based on mass balance to reconstruct historical and daily flow estimates.  The \ac{wrtds} models described below require a matched flow record with the appropriate station to evaluate nutrient trends. Given the complexity of inflows and connectivity of the system, only the inflow estimates from the Sacramento and San Joaquin rivers were used as measures of freshwater influence at each station.  Initial analyses indicated that model fit was not significantly improved with flow estimates from locations closer to each station, nor was model fit improved using lagged times series.  As such, the Sacramento daily flow time series was used to account for flow effects at C3, D19, D26, D28, and MD10, and the San Joaquin time series was used for C10 and P8.  The salinity observations at D4, D6, and D7 in Suisun Bay were used as a more appropriate measure of variation in freshwater balance given the stronger tidal influence at these stations.  Salinity has been used as a tracer of freshwater influence for the application of WRTDS models in tidal waters \cite{Beck15}.

@@

Relationships between flow and observed water quality are complex and can change significantly through space and time\cite{Hirsch10,Zhang16}. These principles have been demonstrated for monitoring data in the Delta region\cite{Jassby08,Novick15,Jabusch16}, suggesting that trend analyses using the observed time series are confounded by flow effects.  Change over for time in observed data could reflect mobilization or dilution effects of flow on concentration rather than an interpretable system response to changes in nutrient sources.  As a proof of concept, \cref{fig:dinc10} demonstrates use of \ac{wrtds} to isolate a flow-normalized time series from the observed \ac{din} data at C10. Raw data are presented in \cref{fig:dinc10}a and the annual results by water year (October through September) from \ac{wrtds} are shown in \cref{fig:dinc10}b. In addition to removing the seasonal component, \cref{fig:dinc10}b shows the flow-normalized component (solid line) independent of the model predictions.  The difference between the two is shown in \cref{fig:dinc10}c such that years with predictions greater or less than the flow-normalized values correspond with long-term trends in flow shown in \cref{fig:dinc10}d.  For example, 1984 is a period of high flow and a large, negative difference between prediction and flow-normalized concentration, suggesting a dilution effect of increased flow on nutrient concentration.  Further, \cref{fig:dinc10}e shows seasonal variation in the relationships of \ac{din} with flow throughout the period of record. Increases in flow (y-axis) were associated with an increase in \ac{din} (colors) for flow values within the observed range.  Seasonal patterns also differed througout the time period with a wider range of \ac{din} within a growing season in the early 2000s relative to the 1980s. % The seasonal shifts are potentially linked to seasonal upwelling linked to ENSO \cite{Cloern07}

@@

# descriptsions for sig changes
zcmp <- select(cmp, -perchg) %>% 
  mutate(ztest = ifelse(ztest < .05, 'sig', 'ns')) %>% 
  spread(dat, ztest) %>% 
  mutate(
    sigtons = ifelse(obs == 'sig' & nrm == 'ns', 1, 0), 
    nstosig = ifelse(obs == 'ns' & nrm == 'sig', 1, 0)
  )

# descriptions for magnitude changes
chgcmp <- select(cmp, -ztest) %>% 
  spread(dat, perchg) %>%
  mutate(
    signobs = sign(obs), 
    signnrm = sign(nrm),
    chgs = ifelse(
      signobs == -1 & nrm < obs, 'moreneg', 
      ifelse(signobs == 1 & nrm > obs, 'morepos',
      ifelse(signobs == 1 & signnrm == -1, 'postoneg', 
      ifelse(signobs == -1 & signnrm == 1, 'negtopos', 
      ifelse(signobs == -1 & signnrm == -1 & nrm > obs, 'lessneg', 
      ifelse(signobs == 1 & signnrm == 1 & nrm < obs, 'lesspos', 
        'somechg')
    )))))
  )
# table of results for shorter sexpr
chgtb <- table(chgcmp$chgs)

@@

A comparison of trends with flow-normalized results from \ac{wrtds} relative to observed data is justified because flow and nutrient concentrations were linked at many of the stations in the study area, similar to \cref{fig:dinc10}.  These comparisons are made relative to changes in the magnitude, significance, and direction of trends, all of which have important implications for decision-making. For all \Sexpr{eng(nrow(zcmp))} trend comparisons in \cref{fig:trndcomp1} (flow-normalized values in \cref{tab:trndsann}) regardless of site, nitrogen analyte, and time period, \Sexpr{eng(sum(zcmp$nstosig))} comparisons had trends that were insignificant with the observed data but significant with flow-normalized results, whereas only \Sexpr{eng(sum(zcmp$sigtons))} trend changed to insignificant. Changes in the magnitude of the estimated percent change per year were also apparent for the flow-normalized trends, such that \Sexpr{eng(sum(chgtb[c('morepos', 'moreneg')]))} comparisons showed an inrease in magnitude (more negative or more positive) and \Sexpr{eng(sum(chgtb[c('lesspos', 'lessneg')]))} had a decrease (less positve or less negative) compared to observed trends.  More importantly, \Sexpr{eng(chgtb['postoneg'])} comparisons showed a trend reversal from positive to negative estimated change and \Sexpr{eng(chgtb['somechg'])} sites went from no change to negative estimated trends for the flow-normalized results. Differences by season in the observed relative to flow-normalized results from \ac{wrtds} were also apparent (\cref{fig:trndcomp2,fig:trndcomp3,tab:trndsbef,tab:trndsaft}).  The most notable changes were an overall decrease in the estimated trend for most sites in the summer and fall seasons for 1996-2013.

Differences in apparent trends underscore the importance of considering flow effects in the interpretation of environmental changes, where flow in the Delta is affected by complex relationships between river inputs, water withdrawals, and tidal exchange with the larger Bay.

Implications of different flow varialbes - refer to \cite{Novick15}, a hybrid approach would be best for complex systems like the delta. Huge effect on trend interpration

@@

@@

### 2017-01-15 sftrends_manu.tex
trndmap gone, almost done with trends text

\title[Water quality change in the upper SFE]{Four decades of water quality change in the upper San Francisco Estuary\footnote{Version: Fri Jan 13 15:40:02 2017 -0600, \href{https://github.com/fawda123/sftrends_manu/commit/f49c0a27672550b75425bf31fd1b561ddf59669e}{f49c0a27672550b75425bf31fd1b561ddf59669e}}}
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\begin{document}
\linenumbers

@@

Relationships between flow and observed water quality are complex and can change significantly through space and time\cite{Hirsch10,Zhang16}. These principles have been demonstrated for monitoring data in the Delta region\cite{Jassby08,Novick15,Jabusch16}, suggesting that trend analyses using the observed time series are confounded by flow effects.  Change over for time in the observed reflect could reflect mobilization or dilution effects of flow on concentration rather than an empirical system response to changes in nutrient sources.  As a proof of concept, \cref{fig:dinc10} demonstrates use of \ac{wrtds} to isolate a flow-normalized time series from the observed data. Raw data are presented in \cref{fig:dinc10}a and the annual results by water year (October through September) from \ac{wrtds} are shown in \cref{fig:dinc10}b. In addition to removing the seasonal component, \cref{fig:dinc10}b shows the flow-normalized component (solid line) independent of the model predictions.  The difference between the two is shown in \cref{fig:dinc10}c such that years with predictions greater or less than the flow-normalized values correspond with long-term trends in flow shown in \cref{fig:dinc10}d.  For example, 1984 is a period of high flow and a large, negative difference between prediction and flow-normalized concentration, suggesting a dilution effect of increased flow on nutrient concentration.  Further, \cref{fig:dinc10}e shows seasonal variation in the relationships of \ac{din} with flow throughout the period of record. Increases in flow (y-axis) were associated with an increase in \ac{din} (colors) for flow values within the observed range.  Seasonal patterns also differed througout the time period with a wider range of \ac{din} within a growing season in the early 2000s relative to the 1980s. % The seasonal shifts are potentially linked to seasonal upwelling linked to ENSO \cite{Cloern07}

A comparison of trends with flow-normalized results from \ac{wrtds} relative to observed data is justified because flow and nutrient concentrations were linked at many of the stations in the study area, similar to \cref{fig:dinc10}.  These comparisons are made relative to changes in the magnitude, significance, and direction of trends, all of which have important implications for decision-making. For all sixty trend comparisons in \cref{fig:trndcomp1} regardless of site, nitrogen analyte, and time period, thirteen comparisons had trends that were insignificant with the observed data but significant with flow-normalized results. The magnitude of all comparisons changed

Overall, the differences in apparent trends underscore the critical importance of considering flow effects in the interpretation of environmental changes, where flow in the Delta is affected by complex relationships between river inputs, water withdrawals, and tidal exchange with the larger Bay.

@@

{\centering \includegraphics[width=0.7\textwidth]{figs/trndmap-1}

\caption[Percent change per year in nitrogen analytes for aggregations by seasons from (a) 1976-1995 and (b) 1996-2013]{Percent change per year in nitrogen analytes for aggregations by seasons from (a) 1976-1995 and (b) 1996-2013.  Changes are based on seasonal Kendall tests of flow-normalized results within each time period. Station names are shown in the top left panel. Station locations have been jittered to reduce overlap. Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.}\label{fig:trndmap}
\end{figure}

>>>>

\title[Water quality change in the upper SFE]{Four decades of water quality change in the upper San Francisco Estuary\footnote{Version: Fri Jan 13 16:19:14 2017 -0600, \href{https://github.com/fawda123/sftrends_manu/commit/66d301f76d51fddfc7d4eeea2ed7b92995d7385f}{66d301f76d51fddfc7d4eeea2ed7b92995d7385f}}}
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\begin{document}
\linenumbers

@@

Relationships between flow and observed water quality are complex and can change significantly through space and time\cite{Hirsch10,Zhang16}. These principles have been demonstrated for monitoring data in the Delta region\cite{Jassby08,Novick15,Jabusch16}, suggesting that trend analyses using the observed time series are confounded by flow effects.  Change over for time in observed data could reflect mobilization or dilution effects of flow on concentration rather than an interpretable system response to changes in nutrient sources.  As a proof of concept, \cref{fig:dinc10} demonstrates use of \ac{wrtds} to isolate a flow-normalized time series from the observed \ac{din} data at C10. Raw data are presented in \cref{fig:dinc10}a and the annual results by water year (October through September) from \ac{wrtds} are shown in \cref{fig:dinc10}b. In addition to removing the seasonal component, \cref{fig:dinc10}b shows the flow-normalized component (solid line) independent of the model predictions.  The difference between the two is shown in \cref{fig:dinc10}c such that years with predictions greater or less than the flow-normalized values correspond with long-term trends in flow shown in \cref{fig:dinc10}d.  For example, 1984 is a period of high flow and a large, negative difference between prediction and flow-normalized concentration, suggesting a dilution effect of increased flow on nutrient concentration.  Further, \cref{fig:dinc10}e shows seasonal variation in the relationships of \ac{din} with flow throughout the period of record. Increases in flow (y-axis) were associated with an increase in \ac{din} (colors) for flow values within the observed range.  Seasonal patterns also differed througout the time period with a wider range of \ac{din} within a growing season in the early 2000s relative to the 1980s. % The seasonal shifts are potentially linked to seasonal upwelling linked to ENSO \cite{Cloern07}

A comparison of trends with flow-normalized results from \ac{wrtds} relative to observed data is justified because flow and nutrient concentrations were linked at many of the stations in the study area, similar to \cref{fig:dinc10}.  These comparisons are made relative to changes in the magnitude, significance, and direction of trends, all of which have important implications for decision-making. For all sixty trend comparisons in \cref{fig:trndcomp1} (flow-normalized values in \cref{tab:trndsann}) regardless of site, nitrogen analyte, and time period, thirteen comparisons had trends that were insignificant with the observed data but significant with flow-normalized results, whereas only one trend changed to insignificant. Changes in the magnitude of the estimated percent change per year were also apparent for the flow-normalized trends, such that fourteen comparisons showed an inrease in magnitude (more negative or more positive) and twenty five had a decrease (less positve or less negative) compared to observed trends.  More importantly, eleven comparisons showed a trend reversal from positive to negative estimated change and ten sites went from no change to negative estimated trends for the flow-normalized results. Differences by season in the observed relative to flow-normalized results from \ac{wrtds} were also apparent (\cref{fig:trndcomp2,fig:trndcomp3,tab:trndsbef,tab:trndsaft}).  The most notable changes were an overall decrease in the estimated trend for most sites in the summer and fall seasons for 1996-2013.

Differences in apparent trends underscore the importance of considering flow effects in the interpretation of environmental changes, where flow in the Delta is affected by complex relationships between river inputs, water withdrawals, and tidal exchange with the larger Bay.

Implications of different flow varialbes - refer to \cite{Novick15}, a hybrid approach would be best for complex systems like the delta. Huge effect on trend interpration

@@
